$\XX$ dubious ways of estimating $\rr$

### 2018-07-03 paper/old_drafts/draft_v4.tex
finally on git

@@

\newcommand{\XX}{\ensuremath{23}} % number of methods
\newcommand{\rr}{\ensuremath{\mathcal{R}_0}}

@@

\section{Introduction}\label{sec:intro}
What has been called ``arguably the most important quantity in the study of epidemics'', $\mathcal{R}_0$, the reproduction number, is an elusive quantity for epidemic modelers \cite{Heesterbeek2002}.  \citet{anderson1992} define $\rr$ as the ``the average number of secondary infections produced when one infected individual is introduced into a host population where everyone is susceptible.''  $\rr$, in some ways, summarizes an entire epidemic.  This quantity is used whether to assess if a disease will break out and what percentage of the population needs to be vaccinated to avoid such an outbreak.  Despite the clear definition of $\rr$, epidemiologists have struggled to create a \textit{de facto} estimator for $\rr$  \citep{hethcote2000}.  The primary issue in estimating $\rr$ is that the quantity is a \textit{property of the model}, meaning that $\rr$ is dependent not only on the usual noise that comes with statistical modeling but also on a variety of assumptions on how researchers assume disease is transmitted through a population.

Typically, disease models are a part of the ``SI-framework'' pioneered by Kermack and McKendrick in the 1920s \citep{getz2006}.  Here, S and I are known as compartments where S stands for ``susceptible'' and I for ``infected.''  The issue arises in how individuals are assumed, either implicitly or explicitly, to travel from the S compartment to the I compartment and any other compartments that have been introduced such as immune, recovered, dead, or exposed, just to name a few.  One of the simplest models in the SI framework is the SIR model where R stands for a ``recovered'' state where individuals are no longer susceptible to the disease \citep{Kermack700}.  Even in this simplified framework, there is much dispute on how $\rr$ is to be estimated arising from assumptions on how populations intermix, if any intervention has been put in place, or whether disease parameters are constant over time.

Other issues of estimating $\rr$ include mathematical versus statistical approaches (i.e. solving for $\rr$ versus estimating $\rr$), how to estimate the quantity in the presence of different available data, and when to estimate $\rr$ in a disease's lifetime.  For instance, if only weekly incidence counts of a disease are available, then we have different methods of estimating $\rr$ available than when we have more data available such as contact networks or the serial interval of a disease.  Another major issue is that since $\rr$ is so difficult to estimate in the first place, the variance of the quantity has largely been glossed over or completely ignored.  This is especially troubling as we are concerned whether $\rr > 1$, which indicates an outbreak of a disease.  A point estimate is simply not enough to determine the truth of this inequality.

In short, $\rr$ is difficult to estimate because it is a property of the model.  This, in turn, makes it difficult to compare $\rr$ across different diseases or even instances of the same disease.  Many of these issues with estimating $\rr$ are summarized and expanded upon by \cite{li2011}.

We review $\XX$ ways to estimate $\rr$ over varying assumptions and in the presence of different sources of data.  The title of this manuscript is in reference to a seminal paper by Moler and Van Loan in 1978 (and its subsequent update in 2003) in which they calculate ``Nineteen dubious ways to compute the exponential of a matrix'' \citep{moler2003}.  We not only review the methods of estimating of $\rr$ but also include recommendations on when to use the methods as well as introducing techniques in which to estimate the variance of estimators.  We compare these estimates, when applicable, and find that $\rr$ is more robust than what was initially presumed.

The rest of this manuscript is organized as follows.  In Section \ref{sec:overview}, we introduce the methods in four broad groups: the survival function, compartment models, exponential growth models, and agent-based models.    In Section \ref{sec:details} we give an overview of each of the methods, what assumptions are used, what data is required, and where the method has been used in the past along with ways in which to calculate the variance for methods, when applicable.  In Section \ref{sec:results}, we give estimates of $\rr$ for the different methods and their variances over a variety of data sets.  Finally, in Section \ref{sec:dis} we discuss our conclusions and recommendations.  In Appendix \ref{sec:term} we include a glossary of terminology used throughout our manuscript.

@@

We chose $\XX$ methods to review in this paper, which does not cover all possible methods to estimate $\rr$.  These methods were chosen as  they represent the type of models typically used to model $\rr$ in addition to their high impact in the field of epidemiology.

In order to organize the methods for estimating \rr, we have chosen four broad classes of estimation methods.  These partitions are meant to be guidelines of distinguishing methods from one another rather than immutable groups.  Many of these methods could have fit in another category or even multiple of the groups chosen.  We have chosen the four classes as application of the survival function and its close relatives, classic compartment models, exponential growth assumptions, and estimation from networks.  We list these point estimate methods below.

@@

  \end{enumerate}
\item Compartment models 
  \begin{enumerate}
  \item SIR model
    \begin{itemize}
    \item Least squares (SIR LS)
    \item Reparametrized least squares (SIR ReLS)
    \item Linear model approximation (SIR LMA)
    \item Linear model approximation, all time points (SIR LMAT)
    \item Max of data (SIR Max)
    \item Smooth max (SIR SMax)
    \item Incidence-prevalence ratio (SIR IPR)
    \item Smooth incidence-prevalence Ratio (SIR SIPR)
    \item Harko's SIR model (HSIR)
    \item Sequential Bayes (SIR SB)
    \end{itemize}
  \item SEIR model (SEIR)
  \item Amplifier model (AMP)
  \item Next generation model (NGM)
  \end{enumerate}
\item Exponential growth 
  \begin{enumerate}
  \item Exponential growth (EG)
  \item Maximum likelihood of secondary infections (MLSI)
  \item Time dependent reproduction number (TDR)
  \item Initial growth rate and final size  (IGRFS)
  \end{enumerate}
\item Estimation from networks 
  \begin{enumerate}
  \item Contact tracing (CT)
  \item Branching process (BP)
  \item Agent-based models (ABM)
  \end{enumerate}
  \end{enumerate}

  Additionally, we describe  interval estimates for the above methods for estimating \rr.

@@

The survival function is intimately tied to the history of \rr, which originates from the field of demography in the late 1800s \citep{dietz1993estimation}.  In demography, we have
\begin{align*}
\rr = \int_0^\infty p(a) \beta(a) da
\end{align*}
where $p(a)$ denotes the probability of a woman surviving age $a$ and $\beta(a)$ the rate of giving birth to a  girl for an individual of age $a$.  This interpretation also explains the name of \rr, which was imported to the field of epidemiology by MacDonald and Smith as reported by \cite{dietz1993estimation}.  In epidemiology, the analogy is that $p(a)$ is like the duration of a disease, and $\beta$ is the transmission rate.

This is the chronologically first and in many ways, the most direct method to calculate $\rr$.  However, this model also seems to be the most difficult to estimate as both $p(a)$ and $\beta (a)$ as $p(a)$ is a distribution and $\beta$ is a function.  Typically, it is also difficult to estimate these parameters before or even during an ongoing disease outbreak.  In turn, $\rr$ can only be estimated after an outbreak, which means it cannot be used to determine ongoing prevention strategies.  However, even having a post-disease estimation of $\rr$ may be useful for future outbreaks of the disease or being able to compare the severity of many diseases.  Another advantage is that this method is very intuitive and is independent of assumptions of how the disease is transmitted from person to person.    We describe four methods attempting to estimate $\rr$ from the survival function:  the survival function alone, direct estimation, and direct parameter estimation.

@@

Let $F(a)$ be the probability that a newly infected individual remains infectious for at least time $a$ (the survival probability, $p(a)$ in the previous definition).  Moreover, let $b(a)$ $(\beta (a)$ in the previous definition), the average number of newly infected individuals that an infectious individual will produce per unit time when infected for total time $a$.  Then $\rr$ is derived in Equation (\ref{eq:r0_survivalfxn}),

@@

This method presents one of the unique difficulties of estimating $\rr$: that of reconciling generations and time.  $\rr$ tells us about the generations of disease, meaning a parent and their offspring but not necessarily when they occur.  This is made difficult as the data we have about a disease is often of a temporal nature.  On the other hand, this method is flexible in that this derivation of $\rr$ is does not place assumptions on how the disease transmits through a population.  A disadvantage is that data about $F(a)$ and especially $b(a)$ is difficult to collect directly.

@@

Direct parameter estimation is a `plug and chug' approach as described by \cite{lipsitch2003} and \cite{dietz1993estimation}, which has an intuitive derivation, similar to that of the survival function.  The intuition is that we can estimate $\rr$ if we know how many others an individual comes into contact, the probability of transmission given contact, and how long the disease lasts within an individual  In this method, we estimate the parameters $k$, the number of contacts per unit time, $b$, the probability of transmission per contact, and $D$ the mean duration of infectiousness.  Then $\rr$ is the product of these parameters, (Equation \ref{eq:dpe}),
\begin{align}\label{eq:dpe}
\rr = kbD.
\end{align}

\cite{lipsitch2003} note that directly estimating $\rr$ in such a way may be difficult due to the available data, as in the previous two methods.  They instead estimate $\rr$ using the mean serial interval, $\omega$, the time from onset of symptoms in an index case to the onset of symptoms in a subsequent case infected by the index patient in addition to the assumption of exponential growth of infected cases.

In the event, that the is estimate may be calculated, uncertainty estimates must be compounded, and to estimate $\rr$ alone, we only use data from the initial outbreak of the disease, so often we have a small sample size, resulting in large confidence sets or credible intervals.

@@

A derivation of $\rr$ similar to the survival function comes from the idea of the von Foerster equations, which were used to calculate the number of cells in a ``deme,'' a constrained area.  The function $n(t, a)$ calculated the number of cells at age $a$ at time $t$ \citep{trucco1965}.  In epidemiology, $n$ represents the number of infections instead of cells. Letting $Y(t, \tau)$ be the number of infections at time $t$ of age $\tau$ and assuming no deaths in the population, the von Foerster equation is
\begin{align*}
  \frac{\partial Y(t, \tau)}{\partial t} +   \frac{\partial Y(t, \tau)}{\partial \tau} = 0,
\end{align*}
where $Y(t,\tau)$ is the number of people at time $t$ who were infected time $\tau$ ago.    \cite{fraser2004factors} derived an expression for $\rr$ from the von Foerster equation and the equation
\begin{align*}
  Y(t,0) &= \int_0^t \beta( \tau) Y(t, \tau) d \tau
\end{align*}
and initial conditions $Y(0,0) = Y_i$ and $Y(t, \tau) = 0$ when $t < \tau$.  Then the reproduction number is Equation \ref{eq:vfr0},
\begin{align}\label{eq:vfr0}
\rr &= \int_0^\infty \beta(\tau) d\tau,
\end{align}
 where $\beta (\tau)$ represents the infectiousness at time $\tau$ since the infection.    Again, this model is more mathematical in nature than practical as $\beta (\tau)$ is difficult to estimate.  To estimate $\rr$, \cite{fraser2004factors} make an approximation that $Y(t, \tau) = K(\tau) \exp (rt)$ and further adapt the method to include prevention strategies, but the main parameter to be estimated is $\beta (\tau)$.  Estimating $\beta (\tau)$ has many of the same issues as trying to estimate $b(a)$ in the survival function in that it is difficult to estimate while the disease is ongoing but is also advantageous in that it does not place assumptions on how the disease is transmitted throughout the population.

@@

A large and likely the most common set of models used estimate $\mathcal{R}$ is that of compartment models (CMs).  As adapted from \cite{daley2001epidemic}, compartment models consist of the following.  CMs model how objects move from one discrete state to another over time and provide a high-level model of how diseases, for instance, may be spread throughout a population.  Within the CM framework, we make four essential assumptions,
\begin{enumerate}
\item The compartments are discrete and have no overlap.
\item The transition of objects into and out of compartments is described by a set of equations.
\item The populations mix homogeneously
  \item The number of objects in each compartment at time $t=0$ is known
  \end{enumerate}

Another common assumption for compartment models, is the law of mass action, a property borrowed from chemistry which says that the mass of the product is proportional to the mass of the reactants.  \cite{anderson1992} adapt the law of mass action to epidemiology saying, ``the net rate of spread of infection is assumed to be proportional to the product of the susceptible people times the density of the infectious individuals.''  Therefore, the flow of objects from one compartment to another may be dependent on the percentage of objects within one or many of the compartments.  The assumption of mass action holds particularly true for disease modelling as typically we assume infections at the next state are proportional to both the susceptible and infected populations at the current state.

@@

In this section, we discuss 14 different CMs used to estimate $\rr$.  Different CMs yield different mathematical derivations of \rr.  That is, for example, an SIR model with non-constant population will yield a different mathematical formula for $\rr$ than the SEIR model (here ``E'' stands for exposed) with non-constant population.  The difference in these two models is that the SEIR model has an extra compartment which in turn effects the estimates of transitions from one compartment to another.  That said, even within the same compartment model framework, we can estimate $\rr$ in a variety of ways.  In fact, we describe 10 methods to estimate $\rr$, within the SIR framework!  Most of these methods are recipes to estimate parameters within the SIR model and may be extended to other CM frameworks such as the SEIR model.  The final method described in this section, the next generation operator (NGM) is a general method to derive \rr from almost any CM.  A particular advantage to CMs compared to survival function methods is that it better fits data that is typically collected about diseases: namely, incidence counts.  CMs are very popular and still in wide-use today.

@@

We begin with the most well known CM, the SIR model.   First introduced to epidemiology by  \cite{Kermack700}, the SIR model governs how individuals transition from \textbf{S}usceptible, \textbf{I}nfected, and \textbf{R}ecovered compartments.  The model is represented graphically in Figure \ref{fig::sir}.  In this model, once an individual is recovered it cannot be infected again.

@@

\end{tikzpicture}
\caption{Depiction of a SIR model.  One can only get the disease once in this model.}\label{fig::sir}
\end{figure}
In this model particular  SIR model, $N$, the number of individuals in constant, as we see no arrows pointing out of the nodes in Figure \ref{fig::sir}. Simple adaptations of the SIR model can include birth and death rates, which will correspondingly change the derivation of $\rr$ (for more information, see Section \ref{sec:ngm}). The parameters  $\beta$ and $\gamma$ have epidemiological meaning.  Here, $\beta$ is the average infection rate and $\gamma$ is the inverse of the average recovery rate.  The model is represented through the ODEs below.  Throughout, we use $X$, $Y$, and $Z$ for stand-ins for the ODEs for $S$, $I$, and $R$, respectively to avoid confusion.
\begin{align}
\systeme{\frac{dX}{dt} = -\frac{\beta XY}{N}, \frac{dY}{dt} = \frac{\beta XY}{N} - \gamma Y, \frac{dZ}{dt} = \gamma Z}. \label{eq::sir}
\end{align}

@@

An epidemic occurs if the rate of change of susceptibles is positive,
\begin{align*}
  \frac{dY}{dt} &> 0 \\
\rightarrow \frac{\beta S I}{N}  - \gamma I &> 0 ,
\end{align*}

that is,  the rate of new infections is greater than the rate of recovery.  So as long as the number of susceptibles is very large, $\frac{S}{N} \approx 1$, then an epidemic will occur if $\rr >1$, as seen in Equation \ref{eq:deriv_sir},
\begin{align}\label{eq:deriv_sir}
  \rr = \frac{\beta}{\gamma} > 1.
  \end{align}
In the ideal set up, we have data of the number of Susceptibles, Infectious, and Recovered individuals at different points over time.  From this we would like to estimate the $\beta$ and $\gamma$ parameters and ultimately $\rr$.

@@

We can use gradient descent or another optimizer to find estimates for these parameters.  If we assume $Y= f(t) + \epsilon_t$ with $\epsilon_t \overset{iid}{\sim} N(0, \sigma^2)$ where $f(t) = \int_0^t \frac{dY}{dt} dt$, then this this method is equivalent to finding the maximum likelihood estimator (MLE) of $\beta$ and $\gamma$.  However, we note that the the $\epsilon_t$ are very likely not independent from one another and so valid inference is still an issue in this framework.  The advantage is that this is a very straightforward way to estimate $\rr$ from the SIR model.

@@

We use the \(\rr\) directly from the above estimation process, which again be done with gradient descent or another optimization process.  Again, we cannot assume $\rr$ has a normal distribution as the errors are likely correlated.  This method is more advantageous in least squares in finding the distribution of $\rr$ as ratios typically give large confidence intervals.

@@

The SIR model has no known closed form solution and so we are already using approximations using numerical integration, albeit small approximations.  In addition to this, data collected from real diseases are typically very noisy in the first place.  \cite{chang2017} posit the idea that an SIR model may be well approximated by a linear model.  We use this idea here and use the ODEs in Equation \eqref{eq::sir} to relate our estimated linear model to that of $\rr$.

Specifically, we fit a linear polynomial of \(t\) with degree \(K= 10\) model to \(X\)
and \(Y\) using least squares to find the coefficients (\(\hat{x}_k\),
\(\hat{y}_k\)),

\begin{align*}
X(t) &= \sum_{k=0}^K x_k t^k\\
{Y}(t) &= \sum_{k=0}^K y_k t^k
\end{align*}

Then, we estimate the derivatives as

\begin{align*}
\hat{X}^\prime(t) &= \sum_{k=1}^K k \hat{x}_k t^{k-1}\\
\hat{Y}^\prime(t) &= \sum_{k=0}^K k \hat{y}_k t^{k-1}
\end{align*}

Then \(\rr\) is calculated as according to the ODEs,
\[\rr = \frac{\hat{X}^\prime(0)}{ \hat{X}^\prime(0) + \hat{Y}^\prime(0)} \cdot \frac{N}{X(0)} \]

Like in the previous cases, we are assuming the true models are

>>>>

\linespread{2}

@@

\newcommand{\XX}{\ensuremath{25}} % number of methods

\newcommand{\xxsir}{\ensuremath{11}} % number of SIR methods

\newcommand{\rr}{\ensuremath{\mathcal{R}_0}}

@@

\section{Introduction}\label{sec:intro}
What has been called ``arguably the most important quantity in the study of epidemics'', $\mathcal{R}_0$, the reproduction number, has been an elusive quantity to estimate for epidemic modelers for nearly a century \citep{Heesterbeek2002}.  Defined by \citet{anderson1992}, $\rr$ (pronounced $R$-naught) is the ``the average number of secondary infections produced when one infected individual is introduced into a host population where everyone is susceptible.''  $\rr$, in some ways, summarizes an entire outbreak of a disease as this quantity is used whether to assess if a disease will break out and what percentage of the population needs to be vaccinated to avoid such an outbreak.  Despite a clear definition of $\rr$, epidemiologists have struggled to create a \textit{de facto} estimator for $\rr$  \citep{hethcote2000}.  The primary issue in estimating $\rr$ is that the quantity is a \textit{property of the model}, meaning that $\rr$ is dependent not only on the usual noise that comes with statistical modeling but also on a variety of assumptions on how researchers assume a disease is transmitted through a population.

Typically, disease models are a part of the ``SI-framework'' pioneered by Kermack and McKendrick in the 1920s \citep{getz2006}.  Here, S and I are known as compartments where S stands for ``susceptible'' and I for ``infected.''  The issue of estimating $\rr$ stems from how individuals are assumed, either implicitly or explicitly, to move from the S compartment to the I compartment and any other compartments that have been introduced such as immune, recovered, dead, or exposed, for example.  One of the simplest models in the SI framework is the SIR model where R stands for a ``recovered'' state where individuals are no longer susceptible to the disease \citep{Kermack700}.  Even in this simplified framework, there is much dispute on how $\rr$ is to be estimated arising from assumptions on how populations intermix, if any intervention has been put in place, or whether disease parameters are constant over time.

Other issues about estimating $\rr$ include mathematical versus statistical approaches (i.e. solving for $\rr$ versus estimating $\rr$), how to estimate the quantity in the presence of types of available data, and when to estimate $\rr$ in a disease's lifetime.  This depends on the ``time constant'' of the disease.  Some diseases like measles progress rapdily, HIV more slowly, prion diseases glacially so.  For instance, if only weekly incidence counts of a disease are available, then we have different methods of estimating $\rr$ available than when we have more data available such as contact networks or the serial interval of a disease.  Another major issue is that since $\rr$ is so difficult to estimate in the first place, the estimated variance of the estimated $\rr$ has largely been glossed over or  ignored.  This is especially troubling because we are concerned whether $\rr > 1$, which indicates an outbreak of a disease.  A point estimate is simply not enough to determine the truth of this inequality.

In short, $\rr$ is an important parameter in epidemiology that is very difficult to estimate.  This, in turn, makes it difficult to compare $\rr$ across different diseases or even different instances of the same disease.  Many of these challenges with estimating $\rr$ and more are summarized and expanded upon by \cite{li2011}.

We review $\XX$ ways to estimate $\rr$ across a number of model assumptions and in the presence of different sources of data.  The title of this manuscript is in reference to a seminal paper by Moler and Van Loan (1978) (and its subsequent update in 2003) in which they calculate ``Nineteen dubious ways to compute the exponential of a matrix'' \citep{moler2003}.  We not only review the methods of estimating of $\rr$ but also include recommendations on when to use the methods as well as introducing techniques which allow estimation the variance.  We compare these estimates, when applicable, and find that $\rr$ is more robust than what was initially presumed.

The rest of this manuscript is organized as follows.  In Section \ref{sec:overview}, we introduce the methods in four broad groups: the survival function, compartment models, exponential growth models, and agent-based models.    In Section \ref{sec:details}, we give an overview of each of the methods, what assumptions are used, what data is required, and where the method has been used in the past along with ways in which to estimate the variance of the methods, when applicable.  In Section \ref{sec:results}, for a number of different data sets, we give estimates of $\rr$ for the different methods and their variances over a variety of data sets.  Finally, in Section \ref{sec:dis} we discuss our conclusions and recommendations.  In Appendix \ref{sec:term} we include a glossary of terminology used throughout our manuscript.

@@

In this manuscript, $\XX$ methods were chosen to give a survey of methods for estimating $\rr$.  This is not exhaustive, but these methods were chosen due to historical importance and impact in the field of epidemiology.

In order to organize the methods for estimating $\rr$, we have chosen four broad classes of estimation methods.  These partitions are meant to be guidelines for distinguishing methods from one another rather than immutable groups.  Many of these methods  fit in  mulltiple of the groups shown.  We have chosen the four groups as application of the survival function and its close relatives, classic compartment models, exponential growth assumptions, and estimation from networks.  We list these point estimate methods below.

@@

  \end{enumerate}
\item Compartment models 
  \begin{enumerate}
  \item SIR model
    \begin{itemize}
    \item Least squares (SIR LS)
    \item Reparametrized least squares (SIR ReLS)
    \item Linear model approximation (SIR LMA)
    \item Linear model approximation, all time points (SIR LMAT)
    \item Max of data (SIR Max)
    \item Smooth max (SIR SMax)
    \item Incidence-prevalence ratio (SIR IPR)
    \item Smooth incidence-prevalence ratio (SIR SIPR)
    \item Log-linear (SIR LL)
    \item Markov chain estimation (SIR MC)
    \item Sequential Bayes (SIR SB)
    \end{itemize}
  \item SEIR model (SEIR)
  \item MSEIR model (MSEIR)
  \item Amplifier model (AMP)
  \item Next generation model (NGM)
  \end{enumerate}
\item Exponential growth 
  \begin{enumerate}
  \item Exponential growth (EG)
  \item Maximum likelihood of secondary infections (MLSI)
  \item Time dependent reproduction number (TDR)
  \item Initial growth rate and final size  (IGRFS)
  \end{enumerate}
\item Estimation from networks 
  \begin{enumerate}
  \item Contact tracing (CT)
  \item Branching process (BP)
  \item Agent-based models (ABM)
  \end{enumerate}
  \end{enumerate}

  Additionally, we describe  interval estimates for the above methods for estimating $\rr$ in order to better answer the question of whether $\rr > 1$, which implies an outbreak of the disease..

@@

The first set of methods are those of hte survival function, which is closely  tied to the history of $\rr$.  The survival function, which originates from the field of demography in the late 1800s, described how many female offspring a woman would produce \citep{dietz1993estimation}.  In demography, we have
\begin{align*}
\rr = \int_0^\infty p(a) \beta(a) da
\end{align*}
where $p(a)$ denotes the probability of a woman surviving to age $a$ and $\beta(a)$ the rate of an individual of age $a$ giving birth to a  girl.   This interpretation also explains the origins of the name of $\rr$, which was imported to the field of epidemiology by MacDonald and Smith \citep{dietz1993estimation}.  Analogously in epidemiology, $p(a)$ is the age of a disease, and $\beta$ is the transmission rate.

This is the chronologically first, and in many ways, the most direct method to estimate $\rr$.  However, this model also seems to be the most difficult to estimate as both $p(a)$ and $\beta (a)$ are random variables, a distribution and funtion, respectively.  In general, it is also difficult to estimate these parameters before or even during an ongoing disease outbreak.  As such, $\rr$ can only be estimated after an outbreak, which means it cannot be used to determine ongoing prevention strategies.  However, having a post-disease estimation of $\rr$ may be useful for preventing future outbreaks of the disease and aids in comparing the severity of many diseases.  Another advantage of this method is it intuitive and independent of assumptions of how the disease is transmitted from person to person.    We describe four methods attempting to estimate $\rr$ from the survival function:  the survival function alone, direct estimation, and direct parameter estimation.

@@

Let $F(a)$ be the probability that a newly infected individual remains infectious for at least time $a$, and let $b(a)$ be the average number of newly infected individuals that an infectious individual will produce per unit time when infected for total time $a$.  Then $\rr$ is derived in Equation (\ref{eq:r0_survivalfxn}),

@@

This method presents one of the unique difficulties of estimating $\rr$: that of reconciling generations and time.  $\rr$ tells us about the generations of disease, meaning an infector and tis infected, but not necessarily when the infections occur.  This reconciliation between generations and time is made difficult as the data we have about a disease is often of a temporal nature.  On the other hand, this method is flexible in that this derivation of $\rr$  does not place assumptions on how the disease transmits through a population.  A disadvantage is that data about $F(a)$ and especially $b(a)$ is difficult to collect directly.

@@

This approach for parameters that are readily available to be estimated, direct parameter estimation, is a `plug and chug' approach as described by \cite{lipsitch2003} and \cite{dietz1993estimation}.  This method has an intuitive derivation, similar to that of the survival function.  We assume that we can estimate $\rr$ if we know how many others an individual comes into contact $(k)$, the probability of transmission given contact $(b)$, and how long the disease lasts within an individual $(D)$.  Then $\rr$ is the product of these parameters, 
\begin{align}\label{eq:dpe}
\rr = kbD.
\end{align}
\cite{lipsitch2003} note that directly estimating $\rr$ in such a way may be difficult due to the available data, as in the previous two methods.  They instead estimate $\rr$ using the mean serial interval, $\omega$, the time from onset of symptoms in an index case to the onset of symptoms in a subsequent case infected by the index patient in addition to the assumption of exponential growth of infected cases, as a proxy for the duration $D$.  Advantages of this method is that it reduces the problem to three parameters that we can conceivably collect data on.  Disadvantages include in the event, that the is estimate may be calculated, uncertainty estimates must be compounded, and to estimate $\rr$ alone, we only use data from the initial outbreak of the disease, so often we have a small sample size, resulting in large confidence sets or credible intervals.

@@

A derivation of $\rr$ similar to the survival function comes from the idea of the von Foerster equations, which were used to calculate the number of cells in a ``deme,'' a constrained area.  The function $Y(t, \tau)$ calculated the number of cells at age $\tau$ at time $t$ \citep{trucco1965}.  In epidemiology, $Y$ represents the number of infections instead of cells. Letting $Y(t, \tau)$ be the number of infections at time $t$ of age $\tau$ and assuming no deaths in the population, the von Foerster equation is
\begin{align*}
  \frac{\partial Y(t, \tau)}{\partial t} +   \frac{\partial Y(t, \tau)}{\partial \tau} = 0,
\end{align*}
where $Y(t,\tau)$ is the number of people at time $t$ who were infected time $\tau$ ago.    \cite{fraser2004factors} derived an expression for $\rr$ from the von Foerster equation, the additional equation
\begin{align*}
  Y(t,0) &= \int_0^t \beta( \tau) Y(t, \tau) d \tau,
\end{align*}
and initial conditions $Y(0,0) = Y_0$ and $Y(t, \tau) = 0$ when $t < \tau$.  Then the reproduction number is Equation \ref{eq:vfr0},
\begin{align}\label{eq:vfr0}
\rr &= \int_0^\infty \beta(\tau) d\tau,
\end{align}
 where $\beta (\tau)$ represents the infectiousness at time $\tau$ since the infection.    Again, this model is more mathematical in nature than practical as $\beta (\tau)$ is difficult to estimate.  To estimate $\rr$, \cite{fraser2004factors} make an approximation that $Y(t, \tau) = K(\tau) \exp (rt)$ and further adapt the method to include prevention strategies, but the main parameter to be estimated is $\beta (\tau)$.  Estimating $\beta (\tau)$ has many of the same issues as trying to estimate $b(a)$ in the survival function in that it is difficult to estimate while the disease is ongoing but is also advantageous in that it does not place assumptions on how the disease is transmitted throughout the population.

@@

A large and the most common set of methods used estimate $\rr$ is that of fitting compartment models (CMs) to disease data.  As described in \cite{daley2001epidemic},  CMs model how objects move from one discrete state to another over time.  These models provide a high-level model of how diseases may be spread throughout a population.  Within the CM framework, we make four essential assumptions,
\begin{enumerate}
\item The compartments are discrete and have no overlap.
\item The transition of objects into and out of compartments is described by a set of known equations, possibly dependent on unknown parameters.
\item The populations mix homogeneously
  \item The number of objects in each compartment at time $t=0$ is known
  \end{enumerate}

Another common assumption for compartment models, is the law of mass action, a property borrowed from chemistry which says that the mass of the product is proportional to the mass of the reactants.  In epidemiology, this means that the proportion of new infections is proportional to the current  number of susceptible and infected \citep{anderson1992}.  Therefore, the flow of objects from one compartment to another may be dependent on the percentage of objects within one or many of the compartments.

@@

In this section, we discuss $\xxsir$ different CM-based methods used to estimate $\rr$.  Different model specifications of CMs yield different mathematical derivations of $\rr$.  That is, for example, an SIR model with non-constant population will yield a different mathematical formula for $\rr$ than the SEIR model (here ``E'' stands for exposed) with non-constant population.  The difference in these two models is that the SEIR model has an extra compartment which in turn effects the estimates of transitions from one compartment to another.  That said, even within the same compartment model framework, we can estimate $\rr$ in a variety of ways using statistical techniques.  In fact, we describe 10 methods to estimate $\rr$, within the SIR framework alone!  Most of these methods are recipes to estimate parameters within the SIR model and may be extended to other CM frameworks such as the SEIR model.  The final method described in this section, the next generation operator (NGM) is a general method to derive $\rr$ from any valid CM.  A particular advantage to CMs compared to survival function methods is that it easier to estimate as data that is typically collected about diseases: namely, incidence counts.

@@

We begin with the most well known CM, the SIR model.   First introduced to epidemiology by  \cite{Kermack700}, the SIR model governs how individuals transition from susceptible, infected, and recovered compartments.  The model is represented graphically in Figure \ref{fig::sir}.

@@

\end{tikzpicture}
\caption{Depiction of a SIR model.  One can only get the disease once in this model.}\label{fig::sir}
\end{figure}
In this model particular  SIR model, $N$, the number of individuals in constant, as we see no arrows pointing out of the nodes in Figure \ref{fig::sir}. Simple adaptations of the SIR model can include birth and death rates, which may correspondingly change the derivation of $\rr$ (for more information, see Section \ref{sec:ngm}). The parameters  $\beta$ and $\gamma$ have epidemiological meaning.  Here, $\beta$ is the average infection rate and $\gamma$ is the inverse of the average recovery rate.  The transition of individuals from one compartment to another is represented through the ODEs below.  Throughout, we use $X$, $Y$, and $Z$ for stand-ins for the ODEs for $S$, $I$, and $R$, respectively to avoid any confusion with $\rr$.
\begin{align}
\systeme{\frac{dX}{dt} = -\frac{\beta XY}{N}, \frac{dY}{dt} = \frac{\beta XY}{N} - \gamma Y, \frac{dZ}{dt} = \gamma Z}. \label{eq::sir}
\end{align}

@@

An epidemic occurs if the rate of change of susceptibles is positive,
\begin{align*}
  \frac{dY}{dt} &> 0 \\
 \frac{\beta S I}{N}  - \gamma I &> 0 ,
\end{align*}

that is,  the rate of new infections is greater than the rate of recovery.  So as long as the number of susceptibles is large, $\frac{S}{N} \approx 1$, then an outbreak will occur if $\rr >1$,
\begin{align}\label{eq:deriv_sir}
  \rr = \frac{\beta}{\gamma} > 1.
  \end{align}
Ideally, we have access data of the number of susceptibles, infecteds, and recovered individuals at different points over time.  From this we would like to estimate the $\beta$ and $\gamma$ parameters and ultimately $\rr$.

@@

We can use gradient descent or another optimizer to find estimates for these parameters.  If we assume $Y= f(t) + \epsilon_t$ with $\epsilon_t \overset{iid}{\sim} N(0, \sigma^2)$ where $f(t) = \int_0^t \frac{dY}{dt} dt$, then this this method is equivalent to finding the maximum likelihood estimator (MLE) of $\beta$ and $\gamma$.  However, we note that the the $\epsilon_t$ are very likely not independent from one another and so valid inference is still an issue in this framework.  Even if it were correct, ratio estimations typically result in large, uninformative confidence intervals.   The advantage is that this is a very straightforward way to estimate $\rr$ from the SIR model.

@@

We use the \(\rr\) directly from the above estimation process, which again be done with gradient descent or another optimization process.  Again, we cannot assume $\rr$ has a normal distribution as the errors are likely correlated.  This method is more advantageous in least squares in finding the distribution of $\rr$ as there is no ratio estimation.

@@

The SIR model has no known closed form solution and so we are already using approximations using numerical integration, albeit small approximations.  In addition to this, data collected from real diseases are typically very noisy in the first place.  \cite{chang2017} discover that an SIR model may be well approximated by a linear model.  We use this idea here and use the ODEs in Equation \eqref{eq::sir} to relate our estimated linear model to that of $\rr$.

Specifically, we fit a linear polynomial of \(t\) with degree \(K= 10\) model to \(X\)
and \(Y\) using least squares to find the coefficients (\(\hat{x}_k\),
\(\hat{y}_k\)),
\begin{align*}
X(t) &= \sum_{k=0}^K x_k t^k\\
{Y}(t) &= \sum_{k=0}^K y_k t^k
\end{align*}
Then, we estimate the derivatives as
\begin{align*}
\hat{X}^\prime(t) &= \sum_{k=1}^K k \hat{x}_k t^{k-1}\\
\hat{Y}^\prime(t) &= \sum_{k=0}^K k \hat{y}_k t^{k-1}
\end{align*}

Following,  \(\rr\) is derived the ODEs,
\[\rr = \frac{\hat{X}^\prime(0)}{ \hat{X}^\prime(0) + \hat{Y}^\prime(0)} \cdot \frac{N}{X(0)}. \]
Like in the previous cases, we are assuming the true models are

\begin{align*}
X(t) &= \sum_{k=0}^K x_k t^k + \epsilon_t\\
  {Y}(t) &= \sum_{k=0}^K y_k t^k + \epsilon_t,
\end{align*}
with $\epsilon_t \overset{iid}{\sim} N(0, \sigma^2)$.  Here, $K=10$ is rather arbitrary and should be selected using some criterion such as AIC in addition with the knowledge of how many data points are available to the user.  Besides this selection of the degree of polynomials to fit, this model is simple to implement and gives very comparable results to using least squares with the SIR model.  Again, we cannot use this model for inference, as our error assumption is violated.  Another disadvantage is that we are using the first time point $t=0$ to estimate $\rr$.  However, as $t=0$ is the boundary of the support of time values, it is likely be a biased estimate as typically seen in linear regression.
